---
title: "Using TRY( ) to diagnose problems writing dplyr and ggplot commands"
author: "Daniel Kaplan"
date: "Data Computing"
output: html_document
---

```{r include=FALSE}
require(DataComputing)
options(TRY_interactive = FALSE)
```

Data wrangling commands are often composed of a sequence of data very operations connected by pipes.  For example, here's a sequence with just two steps, involving the verbs `group_by()` and `summarise()`.

```{r}
YearlyAverages <-
  BabyNames %>% 
  group_by(sex) %>%
  summarise(mean(count))
```

It's easy to make mistakes when composing such commands.  As a teacher, I find that much time is spent trying to find mistakes in students' code.  For students, the time they spend trying to fix mistakes on their own or  waiting for me to get around to helping them is frustrating and, for students who can't find their own mistakes, shakes their confidence.

I want to encourage the beginning student to try to identify the origin of errors and fix things. Although I'm happy to help students, it's better for them not to have to rely on an expert.  Instead, they should become experts themselves. 

To illustrate a mistake that a student (or anyone else!) might make, consider this wrangling command intended to find the yearly average number of births in the US, broken down by sex.  Following current fashion, the student wants to refer to sex with the word "gender."
```{r error=TRUE}
YearlyAverages <- 
  BabyNames %>%
  rename(sex = gender) %>%
  group_by(gender) %>%
  summarise(mean(count))
```

The error message indicates that there's a problem with `gender`, which is very helpful.  `TRY()`, part of the `DataComputing` package, attempts to be more helpful. 

The `TRY()` function examines a data-verb step and checks for common errors.  When it finds an error (or a construction that could be problematic), `TRY()` writes a diagnostic message and asks the user if he or she wants to move on.  It also displays the input data table and the output of the step. (In this document, such interactivity is impossible.  Just the diagnostic messages will be show.)

```{r error = TRUE}
YearlyAverages <- 
  BabyNames %>%
  TRY( rename(sex = gender) ) %>%
  TRY( group_by(gender) )  %>%
  TRY( summarise(mean(count)) )
```

First, note that `TRY()` indicates that there was a problem with the `rename()` step. It also gives a better hint about what the problem is: "Old variable names go to the right of `=`."  This might be enough to help the user fix things, so ...

```{r error = TRUE}
YearlyAverages <- 
  BabyNames %>%
  TRY( rename(gender = sex) ) %>%
  TRY( group_by(gender) )  %>%
  TRY( summarise(mean(count)) )
```

So long as `TRY()` doesn't encounter an error, the result of the command is passed to the next step (if any) in the chain of commands.  Here, both the `rename()` and `group_by()` steps worked find.  The problem identified by `TRY()` in the `substitute()` step isn't fatal, so the chain proceeds to a conclusion.  Nevertheless, `TRY()` points out some possible problems. 

```{r error = TRUE}
YearlyAverages <- 
  BabyNames %>%
  TRY( rename(gender = sex) ) %>%
  TRY( group_by(gender) )  %>%
  TRY( summarise(ave = mean(count, na.rm=TRUE)) )
```

## Examining a step

You can suspend the processing of the command chain by adding a waypoint with the `WAYPOINT()` function. Once suspended, the input to the previous step, and the output from that step are displayed in the `TRY INPUT` and `TRY OUTPUT` tabs of the editor.  This gives you a chance to explore what the step did (and whether it did what you were expecting).  You can have more than one waypoint in a command, so it's best to give each of them a name so you know where you are at.

```{r error = TRUE}
YearlyAverages <- 
  BabyNames %>%
  TRY( rename(gender = sex) ) %>% WAYPOINT(first) %>%
  TRY( group_by(gender) )  %>% WAYPOINT(second) %>%
  TRY( summarise(ave = mean(count, na.rm=TRUE)) )
```

You need to try the above commands interactively to see how this works.

Once you are satisfied with a step, you can delete the following waypoint, or comment it out to deactivate it, like this:

```
TRY( rename(gender = sex) ) %>% ## WAYPOINT(first) %>%
```
    
## Multiple steps in a single TRY( ) 

Putting multiple steps into a single `TRY()` will stop at the first step that displays problems or causes an error.  That might be helpful in screening a long command for problems, but I don't think it's better than arranging one `TRY()` for each step.


```{r error=TRUE, eval=FALSE}
Overall <- 
  KidsFeet %>% 
  TRY(
    group_by(sexx) %>% 
      tally() %>% 
      arrange(nn)
    )
```
 
## TRY( ) with ggplot 
   
`TRY()` has a few diagnostics for `ggplot()` and various geoms. For instance, `TRY()` checks whether aesthetics inside `aes()` are mapped to variables and whether aestheetics outside `aes()` are set to constants.  It also checks the names of aesthetics and whether variables used for mapping are contained in the data table used by a layer.

Since `ggplot()` uses `+` to connect the various commands constructing  layers, facets, themes, etc., you need to pack all of the `ggplot`-related commands into a single `TRY()`.  For example:

```{r error=TRUE}
KidsFeet %>% 
  TRY(
    ggplot(aes(x=width, y=length, shape=sex)) + 
      geom_point(color= 'red') + 
      geom_path(aes(fill = sex))
    )
```
You cannot use `WAYPOINT()` inside of `TRY()`.

If you want to break up the `ggplot` commands into separate `TRY()` statements so that you can set a waypoint, you will have to connect the `TRY()` with the `%>%` pipe, **not** `+`.

```{r error=TRUE, eval=FALSE}
KidsFeet %>% 
  TRY( ggplot(aes(x=width, y=length, shape=sex)) ) %>%
  TRY( geom_point(color= 'red') ) %>% WAYPOINT(dots) %>%
  TRY( geom_path(aes(fill = sex)) )
```
 
Regrettably, when removing the `TRY()` after the statement has been debugged, you will need to pay attention to changing the `%>%` into `+`.  This is prone to mistakes and such mistakes lead to an cryptic error message from `ggplot()`.
    
## Working with Rmd files

It's a good practice to develop your commands within an Rmd file.  When you do this, you will want `TRY()` and `WAYPOINT()` to be interactive when you execute the commands from the Rmd to the console, but you want them non-interactive when compiling the Rmd.

To accomplish this, set the `TRY_interactive` option to `FALSE` in a chunk of the Rmd document:
```
options(TRY_interactive = FALSE)
```

Don't run that chunk in the console.

[In draft: test whether `knitr` is running to disable interactivity?]

## Extending TRY( )

The diagnostics in `TRY()` are heuristics aimed to identify specific, common mistakes.  It's possible to add new heuristics tuned for a specific `dplyr` or `ggplot2` function (or potentially, functions from other packages).  If you identify a common mistake that `TRY()` doesn't handle, send a note describing it to `kaplan@macalester.edu`, or, if you know how, add it into `TRY()` yourself and send a pull request.
